---
title: "Google Data Analytics Capstone Project"
subtitle: "Case Study 1: How Does a Bike-Share Navigate Speedy Success?"
output: html_notebook
---
## Scenario
You work as a junior data analyst in the marketing analyst team of **Cyclistic**, a Chicago-based bike-share firm. The company's future prosperity, according to the director of marketing, hinges on increasing the number of yearly subscribers. As a result, your team is interested in learning how casual riders and annual members use Cyclistic bikes. As a result of these findings,

Your group will devise a new marketing approach to convert casual riders into annual subscribers. However, Cyclistic executives must first be addressed. Your recommendations must be backed up by **compelling data insights and professional data visualizations** in order for them to be approved.



## Business Tasks
- How do annual members and casual riders use Cyclistic bikes differently?
- Why would a non-cyclist purchase a Cyclistic annual membership?
- Design marketing strategies aimed at converting casual riders into annual members.
- what impact digital media might have on their marketing strategies


## Data Source
- Divvy Data

### Setting up my environment
- Importing library `tidyverse` and loading datasets
```{r}
library(tidyverse)

```

### Loading Data sets
```{r}
tripdata_202004 <- read_csv("202004-divvy-tripdata.csv")
tripdata_202005 <- read_csv("202005-divvy-tripdata.csv")
tripdata_202006 <- read_csv("202006-divvy-tripdata.csv")
tripdata_202007 <- read_csv("202007-divvy-tripdata.csv")
tripdata_202008 <- read_csv("202008-divvy-tripdata.csv")
tripdata_202009 <- read_csv("202009-divvy-tripdata.csv")
tripdata_202010 <- read_csv("202010-divvy-tripdata.csv")
tripdata_202011 <- read_csv("202011-divvy-tripdata.csv")
tripdata_202012 <- read_csv("202012-divvy-tripdata.csv")
tripdata_202101 <- read_csv("202101-divvy-tripdata.csv")
tripdata_202102 <- read_csv("202102-divvy-tripdata.csv")
tripdata_202103 <- read_csv("202103-divvy-tripdata.csv")
tripdata_202104 <- read_csv("202104-divvy-tripdata.csv")
tripdata_202105 <- read_csv("202105-divvy-tripdata.csv")
```
### `Combining all the data together`

```{r}
tripdata <- rbind(tripdata_202004,
                  tripdata_202005,
                  tripdata_202006,
                  tripdata_202007,
                  tripdata_202008,
                  tripdata_202009,
                  tripdata_202010,
                  tripdata_202011,
                  tripdata_202012,
                  tripdata_202101,
                  tripdata_202102,
                  tripdata_202103,
                  tripdata_202104,
                  tripdata_202105)

```

### `view the data`
```{r}
glimpse(tripdata)

```
## Data Exploration

```{r}
table(is.na(tripdata))
```
### Columns with missing values
```{r}
colSums(is.na(tripdata))
```

### Missing data would be removed.

```{r}
tripdata_cleaned <- tripdata[complete.cases(tripdata), ]
```

### checking the data again
```{r}
colSums(is.na(tripdata_cleaned))
```

### Data with a "started at" value greater than "ended at" will be removed.
```{r}
tripdata_cleaned <- tripdata_cleaned %>% 
  filter(tripdata_cleaned$started_at < tripdata_cleaned$ended_at)
```


```{r}
glimpse(tripdata_cleaned)
```

- *Create a column called `ride_length` Calculate the length of each ride by subtracting the column `started_at` from the column `ended_at`* 
```{r}
tripdata_cleaned$ride_length <- tripdata_cleaned$ended_at - tripdata_cleaned$started_at

head(tripdata_cleaned$ride_length, 20)

```
### Import lubridate library

```{r}
library(lubridate)
```


*Since they are in numeric term we have to convert it.*
```{r}
tripdata_cleaned$ride_length <- hms::hms(seconds_to_period(tripdata_cleaned$ride_length))
head(tripdata_cleaned$ride_length)

```

### Create a column called “day_of_week,” and calculate the day of the week that each ride started
- wday(x, label = FALSE, abbr = TRUE, ordered = FALSE)



```{r}
tripdata_cleaned <- tripdata_cleaned %>% mutate(day_of_week = wday(started_at, label = F))
```


```{r}
head(tripdata_cleaned)
```

## Calculation
- Calculate the mean of ride_length
- Calculate the max ride_length
- Calculate the mode of day_of_week
```{r}
library(DescTools)
```


```{r}
tripdata_summary <- tripdata_cleaned %>% summarise(ride_length_mean = mean(ride_length), ride_length_max = max(ride_length), mode_day_of_week = Mode(day_of_week))
  
tripdata_summary
```
- Calculate the average ride_length for members and casual riders, group by `member_casual`
```{r}
average_ride_length <- tripdata_cleaned %>% group_by(member_casual) %>% summarise(average_ride_length = mean(ride_length))

average_ride_length
```

- Calculate the average ride_length for users by day_of_week. group by `day_of_week`

```{r}
avg_ride_length <- tripdata_cleaned %>% group_by(day_of_week) %>% summarise(avg_ride_length = mean(ride_length))

avg_ride_length
```
- Calculate the number of rides for users by day_of_week 

```{r}
count_rides <- tripdata_cleaned %>% group_by(day_of_week) %>% summarise(number_rides = n())

count_rides
```

```{r}
tripdata_cleaned %>% select(ride_length,member_casual,day_of_week) %>% group_by(day_of_week,member_casual) %>% summarise(mean = mean(ride_length))
```



```{r}
trip <- tripdata_cleaned %>% mutate(weekday = wday(started_at, label = T)) %>% group_by(member_casual, weekday) %>% summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% arrange(member_casual,weekday)

trip
```
## Visualizing number of rides by type

```{r}
ggplot(data = trip) + aes( x = (weekday), y = number_of_rides, fill = (member_casual))+geom_bar(stat='identity', position = position_dodge(width = 1))+ scale_y_continuous(labels = scales::comma) %>%  labs(title = "Number of Rides by Days and Rider Type",
       subtitle = "Members versus Casual Users") 
```

*Saturday has the highest number of rides for both casual riders and members*

## Average Duration


```{r}
ggplot(data = trip) + aes( x = (weekday), y = average_duration, fill = (member_casual))+geom_bar(stat='identity', position = position_dodge(width = 1))+ scale_y_continuous(labels = scales::comma) %>%  labs(title = "Average Ride Duration by Day and Rider Type",
       subtitle = "Members versus Casual Users") 

```
*This shows that casual riders prefer longer rides*

```{r}
trip_add <- tripdata_cleaned %>% group_by(member_casual,rideable_type) %>% summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% arrange(member_casual,rideable_type)

trip_add
```



```{r}
ggplot(data = trip_add) + aes( x = (rideable_type), y = average_duration, fill = (member_casual))+geom_bar(stat='identity', position = position_dodge(width = 1))+ scale_y_continuous(labels = scales::comma) %>%  labs(title = "Average Ride Duration by  Rideable_type",
       subtitle = "Members versus Casual Users") 
```

*The most popular mode of transportation among riders is docked bicycles.*


```{r}
trip_month <- tripdata_cleaned %>% mutate(month = month(started_at, label = T)) %>% group_by(member_casual, month) %>% summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% arrange(member_casual,month)

trip_month

```


```{r}
ggplot(data = trip_month) + aes( x = (month), y = average_duration, fill = (member_casual))+geom_bar(stat='identity')+ scale_y_continuous(labels = scales::comma) %>%  labs(title = "Average Ride Duration by  Month",
       subtitle = "Members versus Casual Users") 
```
*The longest riders are most active in July.*


```{r}

trip_ride <- tripdata_cleaned %>%  mutate(weekday = wday(started_at, label = T)) %>% group_by(member_casual, weekday) %>% summarise(number_of_rides = n()) %>% arrange(member_casual, weekday)

trip_ride

```


```{r}
ggplot(data = trip_ride) + aes( x = (weekday), y = number_of_rides, fill = (member_casual))+geom_bar(stat='identity')+ scale_y_continuous(labels = scales::comma) %>%  labs(title = "Number of Rides by Days",
       subtitle = "Members versus Casual Users") 
```
*I propose starting a marketing campaign in June to encourage casual riders to our Docked Bikes service on weekends, particularly Saturdays.*



```{r}
trip_station <- tripdata_cleaned %>%  group_by(member_casual,start_station_name ) %>% summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% arrange(member_casual,start_station_name) %>%  arrange(desc(number_of_rides,average_duration))

trip_station

```

We should also pay attention to these stations

- Streeter Sr & Grand Ave
- Lake Shore Dr & Monroe St  
- Millenium Park.
